---
title: "World Data"
author: "Karim Abi Fadel"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(infer)
```


# World Data

```{r, get_data, cache=TRUE}

# load gapminder HIV data
hiv <- read_csv(here::here("data","adults_with_hiv_percent_age_15_49.csv"))
life_expectancy <- read_csv(here::here("data","life_expectancy_years.csv"))

# get World bank data using wbstats
indicators <- c("SP.DYN.TFRT.IN","SE.PRM.NENR", "SH.DYN.MORT", "NY.GDP.PCAP.KD")


library(wbstats)

worldbank_data <- wb_data(country="countries_only", #countries only- no aggregates like Latin America, Europe, etc.
                          indicator = indicators, 
                          start_date = 1960, 
                          end_date = 2016)

# get a dataframe of information regarding countries, indicators, sources, regions, indicator topics, lending types, income levels,  from the World Bank API 
countries <-  wbstats::wb_cachelist$countries

```

```{r}
hiv_tidy <- hiv %>% 
  pivot_longer(cols="1979":"2011", names_to = "date", values_to = "hiv_prev")

life_expectancy_tidy <- life_expectancy %>%
  select(country, c("1960":"2016")) %>% 
  pivot_longer(cols="1960":"2016", names_to = "date", values_to = "life_exp")

# Used outer join over here since we don't want to exclude any values from either data frame

hiv_life_merged <- merge(x = hiv_tidy, y = life_expectancy_tidy, by = c("country","date"), all = TRUE)

# Used outer join for hiv_life_merged and world_bank_data with key country and date. But, used left outer join between that and countries frame to exclude the countries such as Middle East etc. (this is not a country in our other frame)

world_data <- merge(x = merge(x = hiv_life_merged, y = worldbank_data, by = c("country","date"), all = TRUE), y = countries, by = c("iso2c","iso3c"), all.x = TRUE) %>% filter(!is.na(region))
```


## Relationship between HIV prevalence and life expectancy
```{r}
world_data %>% 
  filter(date >= 1979) %>% 
  ggplot(aes(x = hiv_prev, y = life_exp)) +
  geom_smooth() +
  labs(x = "HIV Prevelance in %", y = "Life Expectancy", title = "There is a negative relationship between a population's \nlife expectancy and HIV prevalence") +
  theme_economist()

world_data %>%
  filter(date >= 1979) %>% 
  ggplot(aes(x = hiv_prev, y = life_exp)) +
  geom_point(size = 0.2) +
  geom_smooth() +
  facet_wrap(~region, scales="free") +
  labs(x = "HIV Prevelance in %", y = "Life Expectancy", title = "The relationship is not as prevalent in developed regions\ndue to antiretroviral therapy") +
  theme_clean()
```

## Relationship between fertility rate and GDP per capita
```{r}
world_data %>%
  filter(date >= 1979) %>% 
  ggplot(aes(x = NY.GDP.PCAP.KD, y = SP.DYN.TFRT.IN)) +
  geom_smooth() +
  labs(x = "GDP Per Capita", y = "Fertility Rate", title = "Less developed economies tend to have higher fertility\nrates due to labour demand and lack of contreception") +
  theme_economist()

world_data %>%
  filter(date >= 1979) %>% 
  ggplot(aes(x = NY.GDP.PCAP.KD, y = SP.DYN.TFRT.IN)) +
  geom_smooth() +
  facet_wrap(~region, scales="free") +
  labs(x = "GDP Per Capita", y = "Fertility Rate", title = "This relationship applies less in Middle East\nand Africa since very wealthy families tend to\nhave many children") +
  theme_clean()
```

## Regions that have the most observations with missing HIV data
```{r}
world_data %>%
  group_by(region) %>% 
  summarise(
    countNa = sum(is.na(hiv_prev))
  ) %>% 
  ggplot(aes(x=countNa, y=reorder(region,countNa))) +
  geom_col() +
  labs(x = "Number of missing observations of HIV Prevelance", y="") +
  theme_minimal()
  
```

## How has mortality rate for under 5 changed by region?
```{r}
world_data %>% 
  filter(date >= 1979 & !is.na(SH.DYN.MORT)) %>% 
  group_by(date, region) %>% 
  summarise(
    mean_mortality_rate = mean(SH.DYN.MORT),
  ) %>% 
  ggplot(aes(x = date, y = mean_mortality_rate)) +
  geom_point() +
  facet_wrap(~region, scales = "free") +
  labs(x = "", y = "Child Mortality Rate Per Mille", title = "There is a clear decrease in mortality rates over time (1979-2011)\nin all regions") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

## Top 5 countries that have seen the greatest improvement, as well as those 5 countries where mortality rates have had the least improvement or even deterioration
```{r}
# Since the curves are strictly decreasing, then, the difference of max and min will give the final value minus the initial value
differences <- world_data %>% 
  filter(date >= 1979 & !is.na(SH.DYN.MORT)) %>%
  group_by(region, country.x) %>% 
  summarise(
    difference = max(SH.DYN.MORT) - min(SH.DYN.MORT)
  )

for(val in unique(differences$region)){
  print(differences %>% 
    filter(region == val) %>% 
    arrange(-difference) %>% 
    head(5))

  print(differences %>% 
    filter(region == val) %>% 
    arrange(-difference) %>% 
    tail(5))
}
```


## Relationship between primary school enrollment and fertility rate
```{r}
world_data %>%
  filter(date >= 1979) %>% 
  ggplot(aes(SE.PRM.NENR, SP.DYN.TFRT.IN)) +
  geom_smooth() +
  labs(x = "Primary School Enrollment", y = "Fertility rate", title = "Increased fertility rate can be associated with lower primary\nschool enrollment") +
  theme_clean()
```
