---
title: 'Financial Returns'
author: "Karim Abi Fadel"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: zenburn
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(fivethirtyeight)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(rvest)    # scrape websites
library(purrr)  
library(lubridate) #to handle dates
library(scales)
```


# Returns of financial stocks

Historical data of stock prices is to used calculate returns, and examine the distribution of returns. 

```{r load_nyse_data, message=FALSE, warning=FALSE}
nyse <- read_csv(here::here("data","nyse.csv"))
```

## The table and bar plot show the number of companies per sector, in descending order
```{r companies_per_sector}

companies_sector <- nyse %>%
  group_by(sector) %>%
  count() %>%
  arrange(desc(n))

companies_sector

ggplot(companies_sector, aes(x = n, y = reorder(sector, n))) +
  geom_col() + 
  labs(title = "The finance sector takes the lead",
       subtitle = "Number of companies per sector in NYSE",
       x = "Number of Companies",
       y = "") +
  theme_economist()


```


## Next, the [Dow Jones Industrial Aveareg (DJIA)](https://en.wikipedia.org/wiki/Dow_Jones_Industrial_Average) stocks and their ticker symbols are downloaded. Besides the thirty stocks that make up the DJIA, `SPY` which is an SP500 ETF is downloaded. Let's have some fun with the new data.

```{r, tickers_from_wikipedia}

djia_url <- "https://en.wikipedia.org/wiki/Dow_Jones_Industrial_Average"

#get tables that exist on URL
tables <- djia_url %>% 
  read_html() %>% 
  html_nodes(css="table")


# parse HTML tables into a dataframe called djia. 
# Use purr::map() to create a list of all tables in URL
djia <- map(tables, . %>% 
               html_table(fill=TRUE)%>% 
               clean_names())


# constituents
table1 <- djia[[2]] %>% # the second table on the page contains the ticker symbols
  mutate(date_added = ymd(date_added),
         
         # if a stock is listed on NYSE, its symbol is, e.g., NYSE: MMM
         # We will get prices from yahoo finance which requires just the ticker
         
         # if symbol contains "NYSE*", the * being a wildcard
         # then we jsut drop the first 6 characters in that string
         ticker = ifelse(str_detect(symbol, "NYSE*"),
                          str_sub(symbol,7,11),
                          symbol)
         )

# we need a vector of strings with just the 30 tickers + SPY
tickers <- table1 %>% 
  select(ticker) %>% 
  pull() %>% # pull() gets them as a sting of characters
  c("SPY") # and lets us add SPY, the SP500 ETF
```


```{r get_price_data, message=FALSE, warning=FALSE, cache=TRUE}

myStocks <- tickers %>% 
  tq_get(get  = "stock.prices",
         from = "2000-01-01",
         to   = "2020-08-31") %>%
  group_by(symbol)

```
```{r calculate_returns, message=FALSE, warning=FALSE, cache=TRUE}
#calculate daily returns
myStocks_returns_daily <- myStocks %>%
  tq_transmute(select     = adjusted, 
               mutate_fun = periodReturn, 
               period     = "daily", 
               type       = "log",
               col_rename = "daily_returns",
               cols = c(nested.col))  

#calculate monthly  returns
myStocks_returns_monthly <- myStocks %>%
  tq_transmute(select     = adjusted, 
               mutate_fun = periodReturn, 
               period     = "monthly", 
               type       = "arithmetic",
               col_rename = "monthly_returns",
               cols = c(nested.col)) 

#calculate yearly returns
myStocks_returns_annual <- myStocks %>%
  group_by(symbol) %>%
  tq_transmute(select     = adjusted, 
               mutate_fun = periodReturn, 
               period     = "yearly", 
               type       = "arithmetic",
               col_rename = "yearly_returns",
               cols = c(nested.col))
```

```{r summarise_monthly_returns}

# Monthly returns since 2017-01-01 for each of the stocks and `SPY`; min, max, median, mean, SD.
summary_of_myStocks_returns_monthly <- myStocks_returns_monthly %>%
  filter(date >= as.Date("2017-01-01")) %>% 
  group_by(symbol) %>%
  summarise(min = min(monthly_returns),
            median = median(monthly_returns),
            max = max(monthly_returns),
            mean = mean(monthly_returns),
            std = sqrt(var(monthly_returns)))

summary_of_myStocks_returns_monthly

```

## Here, you can find a density plot of the monthly returns of each of the stocks
```{r density_monthly_returns}
myStocks_returns_monthly %>% 
  filter(symbol != "SPY") %>%
  ggplot(aes(x = monthly_returns)) +
  geom_density() + 
  labs(title = "Density plot of monthly returns", x = "Stock", y = "Monthly returns") +
  facet_wrap(~symbol) +
  theme_gdocs()

```

>From the plot, we can infer that firms with "wider" curves show more volatility. With that in mind, Coca Cola looks like the least risky firm in the data set as its thin peak represents a smaller deviation in stock movements. On the other hand, Dow Inc comes across as the riskiest stock in the dataset due to its wide peak representing a large deviation in stock prices and thus high volatility. 


```{r risk_return_plot}
summary_of_myStocks_returns_monthly %>% 
  filter(symbol != "SPY") %>%
  ggplot(aes(std, mean, label = symbol)) +
  geom_point() + 
  ggrepel::geom_text_repel() +
  labs(title = "KO and DOW are at the extremeties of risk as predicted", subtitle = "Expected monthly returns as a function of standard deviation", x = "Standard Deviation", y = "Expected monthly returns")+
  theme_tq()

```

>We can observe that there is a positive correlation between risk and expected monthly returns; higher risk stocks tend to give higher returns. There are outliers, notably DOW which shows the highest risk between the stocks but relatively low returns. 