---
title: "Climate Change"
author: "Karim Abi Fadel"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(infer)
```



# Climate change and temperature anomalies 


In order to study climate change, we can find data on the *Combined Land-Surface Air and Sea-Surface Water Temperature Anomalies* in the Northern Hemisphere at [NASA's Goddard Institute for Space Studies](https://data.giss.nasa.gov/gistemp). The [tabular data of temperature anomalies can be found here](https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.txt)

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

weather

```

## Reframing the data for smoother analysis
```{r tidyweather}
tidyweather <- weather %>% 
  select(-c("J-D","D-N","DJF","MAM","JJA","SON")) %>% 
  pivot_longer(cols="Jan":"Dec", names_to = "Month", values_to = "delta")

tidyweather

```


## Plotting Information
```{r scatter_plot, eval=TRUE}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), Month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Climate change is real!",
    subtitle = "The graph shows a growing deviation from the 1951 to 1981 reference period",
    x = "Date",
    y = "Delta"
  ) +
  theme_economist()

```

```{r facet_wrap, echo=FALSE}

#Your code goes here...
ggplot(tidyweather, aes(x=year, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Weather anomalies are evident in all months",
    x = "Year",
    y = ""
  ) +
  facet_wrap(~month, scales = 'free', labeller = 'label_value') +
  theme_clean()
```

```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```

```{r density_plot}

ggplot(comparison, aes(x=delta, fill=interval))+
  geom_density(alpha=0.2) +   #density plot with tranparency set to 20%
  theme_bw() +                #theme
  labs (
    title = "Density Plot for Monthly Temperature Anomalies",
    y     = "Density"         #changing y-axis label to sentence case
  )

```

```{r averaging}

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(annual_average_delta = mean(delta, na.rm=TRUE)) 

#plotting the data:
ggplot(average_annual_anomaly, aes(x=Year, y= annual_average_delta))+
  geom_point()+
  
  #Fit the best fit line, using LOESS method
  geom_smooth() +
  
  #change to theme_bw() to have white background + black frame around plot
  theme_bw() +
  labs (
    title = "Climate change has been accelerating exponentially since 1960",
    y     = "Average Annual Delta"
  ) +
  theme_clean()


```


## Confidence Interval for `delta`

[NASA points out on their website](https://earthobservatory.nasa.gov/world-of-change/decadaltemp.php) that 

> A one-degree global change is significant because it takes a vast amount of heat to warm all the oceans, atmosphere, and land by that much. In the past, a one- to two-degree drop was all it took to plunge the Earth into the Little Ice Age.

```{r, calculate_CI_using_formula}
# use formula to construct confidence interval

formula_ci <- comparison %>% 
  filter(interval == "2011-present" & !is.na(delta)) %>% 
  summarise(mean_delta = mean(delta),
            sd_delta = sd(delta),
            count = n(),
            # get t-critical value with (n-1) degrees of freedom
            t_critical = qt(0.975, count-1),
            se_delta = sd(delta)/sqrt(count),
            margin_of_error = t_critical * se_delta,
            delta_low = mean_delta - margin_of_error,
            delta_high = mean_delta + margin_of_error
  ) 

#print out formula_CI
formula_ci
```


```{r, calculate_CI_using_bootstrap}

# use the infer package to construct a 95% confidence interval for delta

set.seed(24)
comparison %>% 
  filter(interval == "2011-present" & !is.na(delta)) %>%
  specify(response = delta) %>% 
  generate(reps = 1000) %>% 
  calculate(stat = "mean") %>% 
  visualise()

```

> The above graph shows that when repeated 1000 times, the data shows an almost normal distribution. and the mean is relatively close to the formula calculation at around 0.96. The more times the test is repeated, the closer the graph moves towards a normal distribution and the closer the mean moves towards the formula calculated value.
